---
title: "Multigene family scan"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
date: "2023-02-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2); theme_set(theme_bw()); theme_update(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))

Parse_blast_wite_bed <- function(input_filename, output_filename) {
  dat = read_tsv(input_filename, col_names = F, show_col_types = FALSE) %>% 
  select(X1, X7, X8, X11, X3) %>% 
  rename(contig = X1, start =X7,  end = X8,  eval = X11, pident = X3) %>% 
  mutate(length = abs(end-start) ) %>% 
  unique() %>% 
  filter(length >150) %>% 
  #filter(contig == "tig00000727")  %>% this was just for checking
  group_by(contig, start) %>% 
  arrange(-length) %>% 
  filter(row_number()==1) %>% ungroup() %>% 
  group_by(contig, end) %>% 
  filter(row_number()==1) %>% ungroup() %>% 
  mutate(strand = ifelse(start - end >0 ,"-", "+"))
  
  dat %>% 
  mutate(new_start = ifelse(strand == "+", start, end) ) %>% 
  mutate(new_end = ifelse(strand =="+", end, start )) %>% 
  select(contig, new_start, new_end, eval, length , strand) %>% 
  arrange(contig, new_start) %>% 
  write_delim(output_filename, col_names = F, delim = "\t")
}

```
#run blast
Against known TSes annotated in the CL brenner genome
```{bash}
blastn -query ../../assembly/nd_v1_polish/genome.nextpolish.fa -subject ./known_gene_families/TranSialidase.fasta -outfmt 6 -out ./blast_wg/TS_Tula_blast.out -max_hsps 100 -perc_identity 85
blastn -query ../../assembly/nd_v1_polish/genome.nextpolish.fa -subject ./known_gene_families/mucin.fasta -outfmt 6 -out ./blast_wg/mucin_Tula_blast.out -max_hsps 100 -perc_identity 85
blastn -query ../../assembly/nd_v1_polish/genome.nextpolish.fa -subject ./known_gene_families/MASP.fasta -outfmt 6 -out ./blast_wg/MASP_Tula_blast.out -max_hsps 100 -perc_identity 85
blastn -query ../../assembly/nd_v1_polish/genome.nextpolish.fa -subject ./known_gene_families/RHS.fasta -outfmt 6 -out ./blast_wg/RHS_Tula_blast.out -max_hsps 100 -perc_identity 85
blastn -query ../../assembly/nd_v1_polish/genome.nextpolish.fa -subject ./known_gene_families/DGF1.fasta -outfmt 6 -out ./blast_wg/DGF1_Tula_blast.out -max_hsps 100 -perc_identity 85
blastn -query ../../assembly/nd_v1_polish/genome.nextpolish.fa -subject ./known_gene_families/GP63.fasta -outfmt 6 -out ./blast_wg/GP63_Tula_blast.out -max_hsps 100 -perc_identity 85


# now against just thecoding revbion
blastn -query ../../call_orfs/cds.fasta -subject ./known_gene_families/TranSialidase.fasta -outfmt 6 -out ./blast_cds/TS_Tula_blast.out -max_hsps 100 -perc_identity 85
blastn -query ../../call_orfs/cds.fasta -subject ./known_gene_families/mucin.fasta -outfmt 6 -out ./blast_cds/mucin_Tula_blast.out -max_hsps 100 -perc_identity 85
blastn -query ../../call_orfs/cds.fasta -subject ./known_gene_families/MASP.fasta -outfmt 6 -out ./blast_cds/MASP_Tula_blast.out -max_hsps 100 -perc_identity 85
blastn -query ../../call_orfs/cds.fasta -subject ./known_gene_families/RHS.fasta -outfmt 6 -out ./blast_cds/RHS_Tula_blast.out -max_hsps 100 -perc_identity 85
blastn -query ../../call_orfs/cds.fasta -subject ./known_gene_families/DGF1.fasta -outfmt 6 -out ./blast_cds/DGF1_Tula_blast.out -max_hsps 100 -perc_identity 85
blastn -query ../../call_orfs/cds.fasta -subject ./known_gene_families/GP63.fasta -outfmt 6 -out ./blast_cds/GP63_Tula_blast.out -max_hsps 100 -perc_identity 85
```



# Parse blast output and output to bed
```{r eval=FALSE, include=FALSE}
## CDS
filelist = list.files(path="../call_orfs/call_surface_genes/blast_cds/", pattern="*blast.out", all.files=TRUE,
    full.names=TRUE)
for (file in filelist) {
  #small filename
  
  fn = str_split(file, "/")
  fn = fn[[c(1,6)]]
  fn_2 = str_split(fn, "_")
  #extra5ct surf prot
  surf_prot_name  = fn_2[[c(1,1)]]
  #extract strain
  strain_name= fn_2[[c(1,2)]]
  #Create df
  Parse_blast_wite_bed(file, paste0("../call_orfs/call_surface_genes/blast_cds/",surf_prot_name, "_",strain_name,"_","intermediate.bed") )
}

## WGS
filelist = list.files(path="../call_orfs/call_surface_genes/blast_wg/", pattern="*blast.out", all.files=TRUE,
    full.names=TRUE)
for (file in filelist) {
  #small filename
  
  fn = str_split(file, "/")
  fn = fn[[c(1,6)]]
  fn_2 = str_split(fn, "_")
  #extra5ct surf prot
  surf_prot_name  = fn_2[[c(1,1)]]
  #extract strain
  strain_name= fn_2[[c(1,2)]]
  #Create df
  Parse_blast_wite_bed(file, paste0("../call_orfs/call_surface_genes/blast_wg/",surf_prot_name, "_",strain_name,"_","intermediate.bed") )
}
```

## Bedtools merge
Must overlap by at least 100!
```{bash eval=FALSE, include=T}
bedtools merge -s -c 6 -o distinct -d 100  -i TS_Tula_intermediate.bed > TS_tula_merge.bed \
bedtools merge -s -c 6 -o distinct -d 100 -i  mucin_Tula_intermediate.bed > mucin_tula_merge.bed  \
bedtools merge -s -c 6 -o distinct -d 100 -i  MASP_Tula_intermediate.bed > MASP_tula_merge.bed \
bedtools merge -s -c 6 -o distinct -d 100 -i  RHS_Tula_intermediate.bed > RHS_tula_merge.bed \
bedtools merge -s -c 6 -o distinct -d 100 -i  DGF1_Tula_intermediate.bed > DGF1_tula_merge.bed \
bedtools merge -s -c 6 -o distinct -d 100 -i  GP63_Tula_intermediate.bed > GP63_tula_merge.bed 

```

## read in data
Let's look at how many of each gene we got we got
```{r}
# whole genome
############THIS IS DIFFERENT THAN CDS. 
filelist = list.files(path="../call_orfs/call_surface_genes/blast_wg/", pattern="*merge.bed", all.files=TRUE,
    full.names=TRUE)
surf_beds = data.frame("contig" = character(),
                       "start" = numeric(),
                       "end" = numeric(),
                       "strand" = character(), 
                       "gene_name" = character(),
                       "strain" = character())
for (file in filelist) {
  #small filename
  
  fn = str_split(file, "/")
  fn = fn[[c(1,6)]]
  fn_2 = str_split(fn, "_")
  #extra5ct surf prot
  surf_prot_name  = fn_2[[c(1,1)]]
  #extract strain
  strain_name= fn_2[[c(1,2)]]
  #Create df
  dat = read_delim(file, col_names = F, show_col_types = FALSE) %>% 
    mutate(gene_name = surf_prot_name ) %>% 
    mutate(strain = strain_name) %>% 
    rename( contig = X1, start = X2, end = X3,strand = X4)
  
  #Add to df
  surf_beds %>% add_row(dat) -> surf_beds
}


surf_beds %>%
  group_by(contig, start,end) %>% unique() %>% 
  mutate(rep = rep(1:n())) %>% 
  mutate(group_sum = sum(rep)) %>% 
  filter(group_sum==1)
  write_tsv("../call_orfs/call_surface_genes/MGF_wg.tsv")
surf_beds %>% 
  group_by(contig, start,end) %>% unique() %>% 
  mutate(rep = rep(1:n())) %>% 
  mutate(group_sum = sum(rep)) %>% 
  filter(group_sum==1)
  write_tsv("../call_orfs/call_surface_genes/MGF_wg.bed", col_names = F)
##########################################
# CDS
############################################
## Bc the start of the cds in the blast result is within the cds region, we gotta get the start and stop to be within context of the rest of the genome
filelist = list.files(path="../call_orfs/call_surface_genes/blast_cds/", pattern="*merge.bed", all.files=TRUE,
    full.names=TRUE)
surf_beds = data.frame("contig" = character(),
                       "start" = numeric(),
                       "end" = numeric(),
                       "strand" = character(), 
                       "gene_name" = character(),
                       "strain" = character())
for (file in filelist) {
  #small filename
  
  fn = str_split(file, "/")
  fn = fn[[c(1,6)]]
  fn_2 = str_split(fn, "_")
  #extra5ct surf prot
  surf_prot_name  = fn_2[[c(1,1)]]
  #extract strain
  strain_name= fn_2[[c(1,2)]]
  #Create df
  dat = read_delim(file, col_names = F, show_col_types = FALSE) %>% 
    mutate(gene_name = surf_prot_name ) %>% 
    mutate(strain = strain_name) %>% 
    rename( contig = X1, start = X2, end = X3,strand = X4)
  
  #Add to df
  surf_beds %>% add_row(dat) -> surf_beds
}
test = Biostrings::readDNAStringSet("../call_orfs/cds.fasta") %>% 
  as.data.frame() %>% rownames_to_column(var = "name") %>% select(name) %>% 
  mutate(contig = str_extract(name , "^.{9}")) %>% 
  mutate(start = str_extract(name , "(?<=:)[:digit:]*?(?=-)") ) %>% 
  mutate(end = str_extract(name, "(?<=-)[:digit:]*" )) %>% 
  mutate(strand = str_extract(name, "(?<=\\()[:punct:]*?(?=\\))")) %>% mutate(strand = ifelse(is.na(strand), "+", "-"))
 
#merge and replace.
dat = left_join(surf_beds %>%
                  rename(name = contig) %>% 
                  select(-start, -end, -strand) ,
                test) %>%
  select(contig, start, end, strand, gene_name) %>% 
  group_by(contig, start,end) %>% unique() %>% 
  mutate(rep = rep(1:n())) %>% 
  mutate(group_sum = sum(rep)) %>% 
  filter(group_sum==1)

dat %>% write_tsv("../call_orfs/call_surface_genes/MGF_cds.tsv")
dat %>% select(contig, start, end, gene_name) %>% write_tsv("../call_orfs/call_surface_genes/MGF_cds.bed", col_names = F)

```
# Graphs 

```{r}
mfg_wgs = read_tsv("../call_orfs/call_surface_genes/MGF_wg.tsv", show_col_types = FALSE)
mfg_cds = read_tsv("../call_orfs/call_surface_genes/MGF_cds.tsv", show_col_types = FALSE)
```


Make graphs - CDS
```{r}

# Size of all surface proteins 
table(mfg_cds$gene_name) %>%
  as.data.frame() %>% 
  rename("gene" = Var1) %>% 
  ggplot(aes(x = gene, y= Freq, fill = gene))+
  geom_bar( position = "dodge", stat= "identity", color = "black")+
  ggtitle("Number of each gene family")

# length for all surface proteins
mfg_cds %>% 
  mutate(length = abs(start - end)) %>% 
  ggplot(aes(x =length)) +
  geom_histogram()+
  facet_wrap(~gene_name, scales = "free") +ggtitle("removing double hits")

#genome percent
mfg_cds %>% 
  mutate(length = abs(start - end)) %>% 
  rename(gene = gene_name) %>% 
  group_by( gene) %>% 
  summarise(length_sum = sum(length))  %>% 
  mutate(percent_genome = (length_sum/48462332 )*100 ) %>% 
  ggplot(aes(x = forcats::fct_rev(fct_reorder(gene, percent_genome)), y= percent_genome, fill = gene))+
  geom_bar( position = "dodge", stat= "identity", color = "black")+
  geom_text(aes(label=signif(percent_genome, digits = 4)), vjust=-.7) +
  ggtitle("Percent of each gene family")


```
Make graphs - wgs
```{r}
table = table(mfg_wgs$strain, mfg_wgs$gene_name) %>% as.data.frame() %>% 
  rename("strain" = Var1, "gene" = Var2)
# Size of all surface proteins 
table %>% 
  filter(strain !="Tula") %>% 
  ggplot(aes(x = gene, y= Freq, fill = strain))+
  geom_bar( position = "dodge", stat= "identity", color = "black")+
  ggtitle("Number of each gene family found in each strain")

table %>% 
  filter(strain !="Tula") %>% 
  ggplot(aes(x = strain, y= Freq, fill = gene))+
  geom_bar( position = "dodge", stat= "identity", color = "black")+
  ggtitle("Number of each gene family found in each strain")

# Percent of genome for all surface proteins

## Make a table of strain genome sizes
strain_genome_size = data.frame(strain = c("Bug", "CLEsm", "Dm","Sylv", "tula", "Y"),
                                size = c(56077138, 33071242, 54160060, 42072610, 48462332, 39698620 ))

mfg_wgs %>% 
  filter(strain !="Tula") %>% 
  mutate(length = abs(start - end)) %>% 
  left_join(strain_genome_size, by = "strain") %>% 
  rename(gene = gene_name) %>% 
  left_join(table, by = c("gene", "strain")) %>% 
  group_by(size, strain, gene, Freq) %>% 
  summarise(length_sum = sum(length))  %>% 
  mutate(percent_genome = (length_sum/size )*100 )  %>% 
  ggplot(aes(x = gene, y= percent_genome, fill = strain))+
  geom_bar( position = "dodge", stat= "identity", color = "black")+
  ggtitle("percent of genome each gene family makes up")
  
mfg_wgs %>%   
  filter(strain !="Tula") %>% 
  mutate(length = abs(start - end)) %>% 
  left_join(strain_genome_size, by = "strain") %>% 
  rename(gene = gene_name) %>% 
  left_join(table, by = c("gene", "strain")) %>% 
  group_by(size, strain, gene, Freq) %>% 
  summarise(length_sum = sum(length))  %>% 
  mutate(percent_genome = (length_sum/size )*100 ) %>% 
  ggplot(aes(x = strain, y= percent_genome, fill = gene))+
  geom_bar( position = "dodge", stat= "identity", color = "black")+
  ggtitle("percent of genome each gene family makes up")
  

# I would guess the shorter ones are all psudo genes, but also possible that the longer ones could also be garbage. how to test this?
# Another thing I can do eventually, is either run hmmer or a blastp on the predicted proteins and decide if that's an easy way to detect psudogenes or not. idk!!!!
# Or looking at the CDSes from Augustus and see how many of them are In CDSes

#much fewer mucins which is interesting, same as previously published work. but also less psuedogenes?

mfg_wgs %>% 
  filter(strain == "tula") %>% 
  mutate(length = abs(start - end)) %>% 
  ggplot(aes(x =length)) +
  geom_histogram()+
  facet_wrap(~gene_name, scales = "free")

## table

mfg_wgs %>% 
  #filter(strain !="tula") %>% 
  mutate(length = abs(start - end)) %>% 
  left_join(strain_genome_size, by = "strain") %>% 
  rename(gene = gene_name) %>% 
  left_join(table, by = c("gene", "strain")) %>% 
  group_by(size, strain, gene, Freq) %>% 
  summarise(length_sum = sum(length))  %>% 
  mutate(percent_genome = (length_sum/size )*100 )  #%>%   filter(strain == "tula") %>% ungroup() %>% summarise(sum = sum(length_sum))

```


# How many of these are peusodo genes?
## bedtools interect
Use the intersect of the coding genes come from augusuts and these
```{r}
# Make a bed file with all surface proteins 

# Coding genes
#bedtools intersect -wa -wb -a ./all_surf.bed -b ../augustus/augustus.bed > surface_coding.bed
# Psuedo genes (-v)
#bedtools intersect -wa -wb -a ./intermediate_files/all.bed -b ../augustus/augustus.bed -v > surface_psuedo.bed
```

## proportion of each
```{r}
surf_coding = read_tsv("../call_orfs/call_surface_genes/surface_coding.bed", col_names = F) %>% 
  filter(X13 == "gene") %>% 
  rename(contig = X1, surf_start = X2, surf_end = X3, surf_strand = X4, gene = X5, orf_start = X7, orf_end =X8, orf_strand = X9, orf_name = X15 ) %>% 
 select(contig,surf_start, surf_end, surf_strand,gene,orf_start,orf_end,orf_strand, orf_name) %>% 
  mutate(type = "coding")

surf_psuedo = read_tsv("../call_orfs/call_surface_genes/surface_psuedo.bed", col_names = F)%>% 
  rename(contig = X1, surf_start = X2, surf_end = X3, surf_strand = X4, gene = X5 )%>% 
 select(contig,surf_start,surf_end,surf_strand,gene) %>% 
  mutate(type = "psuedo")

bind_rows(surf_coding, surf_psuedo) %>% unique() %>% 
  mutate(length =surf_end- surf_start ) %>% 
  ggplot(aes(x= length))+
  geom_histogram(aes(fill = type), color = "black")+
  facet_wrap(~gene, scales = "free")
```


# Find surface genes in ORFs only
## Look only in CDS:
```{bash eval=FALSE, include=T}
blastn -subject ./agustsus whatever -query known_gene_family/TranSialidase.fasta -outfmt 6 -out TS_blast.out evalue 1e-5
```
